---
title: "Introduction to kernels"
author: "SoDa Team"
output: rmarkdown::html_vignette
description: >
  Learn what are kernels and howto best use them to enrich your datasets.
vignette: >
  %\VignetteIndexEntry{Introduction to kernels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, eval = FALSE}
library(osmenrich)
```

## Overview of kernels

Kernels play a center role in data enrichment using `osmenrich`, as they are the tool to weight the features retrieved from OpenStreetMap (OSM). If you do not know OSM yet and the main functionality of this package, please refer to the introductory vignette. 

In this short vignette, you will learn:

* What kernels are

* How to use kernels to enrich your data using `enrich_osm()`

* What the differences between types of kernels are

* How to create your own kernel function


## What are kernels?

Kernels are functions used to weight and "aggregate" features retrieved from OSM. Kernels are used in the main function of the `osmenrich` package, `enrich_osm()`, using the variable `kernel = "<name-kernel>"` which allows you to specify the kernel function that will be used to summarize the features retrieved and it is shown in the example below.


```{r, eval=FALSE}
sf_data_enriched <- sf_data %>%
  enrich_osm(
    # [...]
    kernel = "uniform"
    # [...]
  )
```


As kernels convert distance (or duration, depending on your necessity) vectors to single numbers, one can decide both the weight and the "aggregation" function to use to convert these vectors to a single number. The `osmenrich` package allows you to select both the kernel and the aggregation function used to reduce these vectors as follows:

* `kernel = ` can take the following inputs: `"uniform"` if, `"gaussian"` if, `"parabola"` if. This variable is set to `"uniform"` by default.

* `reduce_fun = ` can be any aggregation (or reduction) function such as `sum`, `mean`, `median`, etc. This variable is set to `sum` by default.


Thus, when choosing the kernel for your enrichment call, you usually will set these two variables as follows below

```{r, eval=FALSE}
sf_data_enriched <- sf_data %>%
  enrich_osm(
    # [...]
    kernel = "uniform"
    reduce_fun = sum
    # [...]
  )
```

### Kernels

To make you understand the difference in the way the kernel weights the feature points retrieved in OSM, we plot the three main functions available in `osmenrich`. As you can see in the plot below, each kernels weights differently the features retrieved from OSM depending on the distance from the point of reference around which you are retrieving the points.

```{r, echo=FALSE, message=FALSE}
# Load the required libraries to plot
library(tidyverse)
library(firatheme)
```

```{r, message=FALSE, echo=FALSE, fig.width=7}
# 2d kernel plot
tibble(x = seq(-120, 120, length.out = 1000)) %>% 
  mutate(
    gaussian = osmenrich::kernel_gaussian(x, 100, I),
    parabola = osmenrich::kernel_parabola(x, 100, I),
    uniform  = osmenrich::kernel_uniform(x, 100, I),
  ) %>% 
  pivot_longer(cols = -x, names_to = "kernel", values_to = "weight") %>% 
  ggplot(aes(x = x, y = weight)) +
  geom_line() +
  theme_fira() +
  scale_colour_fira() +
  facet_wrap(~kernel, scales = "free") +
  labs(x = "Distance from reference", y = "Weight")
```

* The `gaussian` kernel weights the features retrieved using the classical "bell curve" shape. The features on the border of the retrieved space are weighted zero.

* The `parabola` kernel

* Finally, the `uniform` kernel gives the same weight to all the features retrieved. 

### Aggregation functions

As stated before, aggregation functions are fundamental when choosing how to aggregate data retrived from OSM. While kernel helps you weight the points retrieved by their distances from the reference points, the aggregation function performs the task of reducing the weighted points to a single value. 

These aggregation functions, called `reduce_fun` in `enrich_osm()`, take any standard R function. Therefore, if you just want to sum all the weighted points retrieved you will use `reduce_fun = sum`; if you instead want to take the mean of the retrieved points you will use `reduce_fun = mean`; for the median `reduce_fun = median` and so on...

Most of the time you will find yourself wanting to just `sum` the (un)weighted points retrieved, but keep in mind that you can also aggregate them using different functions.

## How to use kernels

In the examples above, you were introduced to the syntax needed to choose both the kernel and the aggregation (`reduce_fun`) function. Now, you will go through a basic example to see the differences between kernels.

Borrowing from the base example in our introductory vignette, you first create an example dataset to enrich `sf_example`.

```{r, message=FALSE}
# Load packages used in this example, install if needed
library(tidyverse)
library(sf)
library(osmenrich)

# Create an example dataset to enrich
sf_example <-
  tribble(
    ~person,  ~lat,   ~lon,
    "Alice",  52.12,  5.09,
    "Bob",    52.13,  5.08,
  ) %>%
  sf::st_as_sf(
    coords = c("lon", "lat"),
    crs = 4326
  )
```

Once the `sf` dataset is created, you can enrich it using the base syntax:

```{r, eval=FALSE}
sf_enriched_uniform <- sf_example %>%
  enrich_osm(
    name = "n_waste_baskets",
    key = "amenity",
    value = "waste_basket",
    r = 500
  )
```

Which, behind the scenes, sets automatically the kernel variable to `"uniform"` and the aggregation function to `sum`, as seen below.

```{r, eval=FALSE}
sf_enriched_uniform <- sf_example %>%
  enrich_osm(
    name = "n_waste_baskets",
    key = "amenity",
    value = "waste_basket",
    r = 500,
    # Default parameters now written out for clarity
    kernel = "uniform",
    reduce_fun = sum
  )

sf_enriched_uniform
```

### Changing the kernel

Taking as starting point the previous example, you can modify either the `"kernel"` variable or the `reduce_fun` variable to check the difference in the resulting aggregations.

Modify the `"kernel"` variable to `"gaussian"`, keeping the aggregation function to `sum`, and print the enriched dataset.

```{r, eval=FALSE}
sf_enriched_gaussian <- sf_example %>%
  enrich_osm(
    name = "n_waste_baskets",
    key = "amenity",
    value = "waste_basket",
    r = 500,
    # Default parameters now written out for clarity
    kernel = "gaussian",
    reduce_fun = sum
  )

sf_enriched_gaussian
```

As you can see, the

### Plotting kernels


```{r, message=FALSE, echo=FALSE, fig.width=7}
library(ggspatial)
library(ggplot2)

ggplot() +
  annotation_map_tile(zoomin = 0) +
  geom_sf(data = sf_example_enriched) +
  scale_fill_viridis_c() +
  annotation_scale(location = "tr") + 
  labs(title = "Position of the two people", fill = "") +
  theme_fira() +
  theme(axis.text.x = element_text(angle = 90))
```

### Changing the aggregation function

XXX

## Writing your own kernel

XXX



